Staged Abstract Interpreters

### 2019-02-22 paper/10_intro.tex
intro

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to compilers; by abstraction, we can
refactor interpreters to static analyzers. As an intellectual quest, it is
natural to ask --- starting from an interpreter, \textbf{can we derive a sound
abstract interpreter but it works like a compiler and generates fast analysis
code?} Similar to the concrete interpreter, for an abstract interpreter
$\widehat{\texttt{eval}}$ that takes a program, an abstract environment and
returns abstract values, shown as the following:

\begin{lstlisting}
def $\widehat{\texttt{eval}}$(e: Expr)($\rho$: $\widehat{\texttt{Env}}$): $\widehat{\texttt{Value}}$ = ...
\end{lstlisting}

Given a program $e_0$ statically, We would like to have a specialized abstract
interpreter $\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, but leave the abstract environment $\rho$ dynamic.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=4em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}

>>>>

Now we have seen two orthogonal transformations of interpreters: by
speclization, we can refactor interpreters to code generators; by abstraction,
we can refactor interpreters to static analyzers. As an intellectual quest, it
is natural to ask --- starting from an interpreter, \textbf{can we derive a
sound abstract interpreter but it works like a code generator which produces fast
analysis code?} Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
repsect to a program $e_0$ and produce an code generator
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the staged one leaves the
abstract environment $\widehat{\rho}$ dynamic.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}
